A preliminary constraint grammar for Russian

### 2015-04-30 2015-nodalida-ruscg/ruscg.tex
adding reviewer comments as marginpars
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60197 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
\usepackage{tipa} %allows IPA symbols
\usepackage{linguex}
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage[small,bf]{caption}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\begin{document}
\maketitle
\begin{abstract}
 This paper presents preliminary work on a constraint
 grammar based disambiguator for Russian. Russian is
 a Slavic language with a high degree of both in-category
 and out-category homonymy in the inflectional system.
 The pipeline consists of a finite-state morphological
 analyser and constraint grammar. The constraint 
 grammar is tuned to be high recall at the expense of 
 low precision.
 \rr{more precise results here}
\end{abstract}

@@

This paper presents a preliminary constraint grammar for Russian. The principle 
objective of the constraint grammar is to produce a high recall grammar to serve
as input into other natural language processing tasks.

@@

We identify three different types of morphosyntactic ambiguity: 
intraparadigmatic, morphosyntactically incongruent, and morphosyntactically 
congruent. The following examples make use of stress ambiguity to illustrate 
each kind of ambiguity.\footnote{A recent study by \newcite{reynolds.tyers-15} 
found that about 7.5\% of 
morphosyntactic ambiguity in a corpus of Russian resulted in stress placement 
ambiguity.} \emph{Intraparadigmatic} ambiguity refers to homographic 
wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

@@

\needspace{6\baselineskip} % keeps example all together (no orphan line) 
\ex. Morphosyntactically congruent homographs \label{ex:MScongruent}
\a. \rus{з\'{a}мок} \emph{z\'{a}mok} `castle.\textsubscript{SG-NOM}'\\
	\rus{зам\'{о}к} \emph{zam\'{o}k} `lock.\textsubscript{SG-NOM}'
\b. \rus{з\'{a}мков} \emph{z\'{a}mkov} `castle.\textsubscript{PL-GEN}'\\
	\rus{замк\'{о}в} \emph{zamk\'{o}v} `lock.\textsubscript{PL-GEN}'
\c. ...\\
	...

@@

Table~\ref{table:ambiguity} shows that most morphosyntactic ambiguity in 
unrestricted Russian text is rooted in intraparadigmatic and morphosyntactically 
incongruent ambiguity. Detailed part-of-speech tagging with morphosyntactic 
analysis can help disambiguate these forms. However, morphosyntactically congruent stress 
ambiguity cannot be disambiguated by means of detailed part-of-speech tagging, 
and since congruent stress ambiguity represents only a small percentage of 
ambiguous wordforms, we leave it to future work.

@@

Due to the lack of a hand-annotated corpus compatible with our morphological
analyser, we adopted a slightly modified technique. We picked a random text
from the Russian Wikipedia, ran it through the morphological analyser, wrote 
rules, and then ran the rules on the whole Wikipedia corpus. For each rule,
we collected around 100 example applications and checked them. If a rule
selected the appropriate reading in all cases, we included it in the \emph{safe}
rule set, if it removed an appropriate reading in less then three cases, 
then we included it in the \emph{safe heuristic} rule set. Otherwise we either
discarded the rule or included it in the heuristic rule set.

@@

\subsection{Intrinsic}
%FRAN

@@

% unknown words from test corpus, categorisation

@@

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{graphics/learning-curve.pdf}
  \caption{Learning curve for three taggers, \texttt{hunpos} with no lexicon, 
    \texttt{hunpos} with a lexicon, and \texttt{hunpos} with a lexicon and the 
    Russian constraint grammar in a voting set up.}
  \label{fig:curve}
\end{figure}

\section{Future work}
\label{sec:future}
% ROB
% FRAN
% func analysis
% dep analysis

@@

This paper has presented a preliminary constraint grammar for Russian. The constraint grammar 
is high recall and improves the performance over a simple HMM-based tagger. It also shows 
state-of-the-art performance for the stress-placement task.

>>>>

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
\usepackage{tipa} %allows IPA symbols
\usepackage{linguex}
\usepackage[multiple]{footmisc} % separates consecutive footnotes with comma
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage[small,bf]{caption}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

@@

\begin{document}
\maketitle
\begin{abstract}
 This paper presents preliminary work on a constraint
 grammar based disambiguator for Russian. Russian is
 a Slavic language with a high degree of both in-category
 and out-category homonymy in the inflectional system.
 The pipeline consists of a finite-state morphological
 analyser and constraint grammar. The constraint 
 grammar is tuned to be high recall at the expense of 
 low precision.
 \rr{add results here}
\end{abstract}

@@

This paper presents a preliminary constraint grammar for Russian. The main 
objective of the constraint grammar is to produce a high recall grammar to serve
as input into other natural language processing tasks.
\rr{R3: What is the rationale for choosing high recall at the expense of low precision?
Was it a deliberate choice or a consequence of limited development time? In my
experience, this strategy is good for corpus queries in linguistic research
(fewer relevant constructions are missed) but might be problematic for language
technology applications, which normally require full disambiguation. Since it is
stated in the introduction that the principal objective of the grammar is to
produce input to other language processing tasks, this question is especially
relevant and warrants some discussion.}

@@

We identify three different types of morphosyntactic ambiguity: 
intraparadigmatic, morphosyntactically incongruent, and morphosyntactically 
congruent. The following examples make use of stress ambiguity to illustrate 
each kind of ambiguity.\footnote{Written standard Russian does not typically
indicate stress position.}\footnote{A recent study by \newcite{reynolds.tyers-15} 
found that about 7.5\% of 
morphosyntactic ambiguity in a corpus of Russian resulted in stress placement 
ambiguity.} \emph{Intraparadigmatic} ambiguity refers to homographic 
wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

@@

\needspace{6\baselineskip} % keeps example all together (no orphan line) 
\ex. Morphosyntactically congruent homographs \label{ex:MScongruent}
\a. \rus{з\'{a}мок} \emph{z\'{a}mok} `castle.\textsubscript{SG-NOM}'\\
	\rus{зам\'{о}к} \emph{zam\'{o}k} `lock.\textsubscript{SG-NOM}'
\b. \rus{з\'{a}мков} \emph{z\'{a}mkov} `castle.\textsubscript{PL-GEN}'\\
	\rus{замк\'{о}в} \emph{zamk\'{o}v} `lock.\textsubscript{PL-GEN}'
\c. etc. \ldots\\ \rr{is this clear that the paradigm continues, but we don't show it?}

@@

Table~\ref{table:ambiguity} shows that most morphosyntactic ambiguity in 
unrestricted Russian text is rooted in intraparadigmatic and morphosyntactically 
incongruent ambiguity. Detailed part-of-speech tagging with morphosyntactic 
analysis can help disambiguate these forms. However, morphosyntactically congruent
ambiguity cannot be disambiguated by means of detailed part-of-speech tagging, 
and since congruent stress ambiguity represents only a small percentage of 
ambiguous wordforms, we leave it to future work.

@@

Due to the lack of a hand-annotated corpus compatible with our morphological
analyser, we adopted a slightly modified technique. We picked a random text
from the Russian Wikipedia, ran it through the morphological analyser, wrote
\rr{R1: Is ru-wikipedia representative of Russian texts on the whole?}
rules, and then ran the rules on the whole Wikipedia corpus. For each rule,
we collected around 100 example applications and checked them. If a rule
selected the appropriate reading in all cases, we included it in the \emph{safe}
rule set, if it removed an appropriate reading in less then three cases, 
then we included it in the \emph{safe heuristic} rule set. Otherwise we either
discarded the rule or included it in the heuristic rule set.

@@

\subsection{Intrinsic}
%FRAN
\rr{R1 does not understand "intrinsic/extrinsic" (and neither do I) Should we rename them to be qualitative/quantitative?}
\rr{R1: Add some text introducing intrinsic evaluation. What is this section doing?}

@@

\rr{R1: Tables 3 and 4 are not discussed enough.}
% unknown words from test corpus, categorisation

@@

    \rr{R2: The first two rules in 6.2 are a bit odd - the effect would be exactly the same without the first context bracket. Did you mean 0C instead of 0?}

@@

\rr{R1: Note that Johannessen et al (2012) and Lynum et al (2011) have written two articles about building a statistical tagger on top of a cg grammar.

Johannessen, Janne Bond; Hagen, Kristin; Lynum, André; Nøklestad, Anders. 2012. OBT+stat: A combined rule-based and statistical tagger. In Andersen, Gisle (ed.). Exploring Newspaper Language: Using the web to create and investigate a large corpus of modern Norwegian, s. 51–66.

Lynum, Andre; Hagen, Kristin; Johannessen, Janne Bondi; Nøklestad, Anders. OBT+Stat: Evaluation of a combined CG and statistical tagger. In NEALT Proceedings Series 2011 Vol. 14.}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{graphics/learning-curve.pdf}
  \caption{Learning curve for three taggers, \texttt{hunpos} with no lexicon, 
    \texttt{hunpos} with a lexicon, and \texttt{hunpos} with a lexicon and the 
    Russian constraint grammar in a voting set up.}
  \label{fig:curve}
\end{figure}
\rr{R1: Figure 3 is small and hard to read. Shapes indistinguishable.}
\rr{R2: If I interpret Fig. 3 correctly, the HMM system alone is much more dependent on training corpus size than the voting set-up, where accuracies actually appear to become asymptotic already with 8-10,000 words in the training corpus, indicating that further improvement depends more on the CG than the HMM-component.}
\section{Future work}
\label{sec:future}
% ROB
% FRAN
% func analysis
% dep analysis

@@

This paper has presented a preliminary constraint grammar for Russian. The constraint grammar 
is high recall and improves the performance over a simple HMM-based tagger. It also shows 
state-of-the-art performance for the stress-placement task.
\rr{R2: I particularly like the idea of systematically assigning rules to heuristicity sections based on performance thresholds derived from annotation-inspection of a non-gold corpus. Another novel idea (at least in terms of actual implementation) is to go for 100\% recall and add a hybridized HMM-tagger in order to improve precision.}
